Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-06-04 main.tex
Combined second stats table.

In the early years of developments automatic control, \citep{Whipple1899}
remarkably realized and showed that roll motion feedback can stabilize the a
bicycle. Much later, automatic roll stabilization of single track vehicles
began in earnest after predictive motorcycle models were developed and refined
in throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's
work and seems to be the first to attempt to robotically stabilize a small
motorbike with a controlled inverted pendulum that mimicked rider lean, but he
was not successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} sucessfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
propose how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities, which are not prominent in bicycles at low
speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
stability and their vehicle was fully automatic, no human rider was involved.
Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to change the dynamics,
up to some limits, of the human controlled plant. In our prior study, we
demonstrate reduced motion during distractions~\cite{Alizadehsaravi2023} due to
the balance assist but conversely researchers with a similar vehicle recently
showed rider disatisfction with the stabilization~\cite{Hanakam2023}.

@@

\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
\end{align}

The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
functions of the equilbirum forward speed \(v\) and  are otherwise populated
with expressions that are functions of the geometric and inertial parameters of
the nonholonomic mulitbody system made up of four rigid bodies: two wheels,
front frame, and rear frame.

@@

\todo{Combine tables 3 and 4 into one}

\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates at
    6~\unit{\kilo\meter\per\hour} and gain $k=10$. Multiplicative change in odds of predictor variables at 6
  \unit{\kilo\meter\per\hour} and gain $k=10$ based on frequentist
         single-level logistic regression.}
  \label{table:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & Estimate & Standard error & p-value & MCIO & 2.5\% & 97.5\% \\
    \midrule
    Intercept & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse & 1.69 & 0.27 & 0.00 & 5.40 & 3.18 & 9.16 \\
    Perturbation order & -0.77 & 0.22 & 0.00 & 0.46 & 0.30 & 0.72 \\
    Balance assist state & -0.64 & 0.27 & 0.02 & 0.53 & 0.31 & 0.89 \\
    Roll angle & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance-assist on $\times$ roll angle & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance-assist on $\times$ steer angle & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance-assist on $\times$ angular impulse & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance-assist on $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\todo{Combine tables 4 and 5 into one}

\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates at
    10~\unit{\kilo\meter\per\hour} and gain $k=8$.}
  \label{table:freq-coefs}
  \begin{tabular}{lrrr}
    \toprule
    Variable                                      & Estimate          & Standard error          & p-value \\
    \midrule
    Intercept                                     & -0.24             & 0.16                    & 0.13    \\
    Angular impulse                               & 2.39              & 0.29                    & 0.00    \\
    Perturbation order                            & -1.16             & 0.21                    & 0.00    \\
    Balance-assist on                             & -0.44             & 0.24                    & 0.07    \\
    Roll angle                                    & 0.27              & 0.22                    & 0.22    \\
    Steer angle                                   & -0.37             & 0.24                    & 0.12    \\
    Balance-assist on $\times$ roll angle         & -0.61             & 0.34                    & 0.08    \\
    Balance-assist on $\times$ steer angle        & 0.56              & 0.35                    & 0.11    \\
    Balance-assist on $\times$ angular impulse    & 0.46              & 0.44                    & 0.29    \\
    Balance-assist on $\times$ perturbation order & -0.37             & 0.32                    & 0.25    \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}
  \centering
  \caption{Multiplicative change in odds of predictor variables at 10
    \unit{\kilo\meter\per\hour} based on frequentist single-level logistic
    regression.}
  \label{table:freq-change-in-odds}
  \begin{tabular}{lllll}
    \toprule
    Variable & Multiplicative change in odds & 2.5\% & 97.5\% & p-value \\ \hline
    \midrule
    Intercept & 0.78 & 0.57 & 1.07 & 0.12 \\
    Angular impulse & 10.92 & 6.23 & 19.13 & 0.00 \\
    Perturbation order & 0.31 & 0.21 & 0.48 & 0.00 \\
    Balance-assist on & 0.64 & 0.40 & 1.03 & 0.07 \\
    Roll angle & 1.31 & 0.85 & 2.04 & 0.22 \\
    Steer angle & 0.69 & 0.43 & 1.10 & 0.12 \\
    Balance-assist on $\times$ roll angle & 0.54 & 0.28 & 1.07 & 0.08 \\
    Balance-assist on $\times$ steer angle & 1.76 & 0.88 & 3.50 & 0.11 \\
    Balance-assist on $\times$ angular impulse & 1.59 & 0.68 & 3.74 & 0.29 \\
    Balance-assist on $\times$ perturbation order & 0.69 & 0.37 & 1.30 & 0.25 \\
    \bottomrule
  \end{tabular}
\end{table}

>>>>

In the early years of developments in automatic control, \citep{Whipple1899}
remarkably realized and showed that roll motion feedback can stabilize the a
bicycle. Much later, automatic roll stabilization of single track vehicles
began in earnest after predictive motorcycle models were developed and refined
in throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's
work and seems to be the first to attempt to robotically stabilize a small
motorbike with a controlled inverted pendulum that mimicked rider lean, but he
was not successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} sucessfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
proposed how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities, somethign not prominent in bicycles at low
speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
stability and their vehicle was fully automatic, no human rider was involved.
Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to change the dynamics,
up to some limits, of the human controlled plant. In our prior
study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
distractions due to the balance assist but contrarily researchers with a
similar vehicle recently showed rider disatisfction with the
stabilization~\cite{Hanakam2023}.

@@

\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
functions of the equilbirum forward speed \(v\) and  are otherwise populated
with expressions that are functions of the geometric and inertial parameters of
the nonholonomic mulitbody system made up of four rigid bodies: two wheels,
front frame, and rear frame.

@@

\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates at
    6~\unit{\kilo\meter\per\hour} and gain $k=10$. Multiplicative change in
    odds of predictor variables at 6 \unit{\kilo\meter\per\hour} and gain
    $k=10$ based on frequentist single-level logistic regression.}
  \label{table:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & Estimate & Standard error & \(p\)-value & MCiO & 2.5\% & 97.5\% \\
    \midrule
    Intercept & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse & 1.69 & 0.27 & 0.00 & 5.40 & 3.18 & 9.16 \\
    Perturbation order & -0.77 & 0.22 & 0.00 & 0.46 & 0.30 & 0.72 \\
    Balance assist state & -0.64 & 0.27 & 0.02 & 0.53 & 0.31 & 0.89 \\
    Roll angle & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance-assist on $\times$ roll angle & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance-assist on $\times$ steer angle & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance-assist on $\times$ angular impulse & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance-assist on $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates at
    10~\unit{\kilo\meter\per\hour} and gain $k=8$.  Multiplicative change in
    odds of predictor variables at 10 \unit{\kilo\meter\per\hour} based on
    frequentist single-level logistic regression.}
  \label{table:freq-coefs}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & Estimate & Standard error & \(p\)-value & MCiO & 2.5\% & 97.5\% \\
    \midrule
    Intercept                                     & -0.24 & 0.16 & 0.13 & 0.78 & 0.57 & 1.07 \\
    Angular impulse                               & 2.39  & 0.29 & 0.00 & 10.92 & 6.23 & 19.13 \\
    Perturbation order                            & -1.16 & 0.21 & 0.00 & 0.31 & 0.21 & 0.48 \\
    Balance-assist on                             & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
    Roll angle                                    & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
    Steer angle                                   & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
    Balance-assist on $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
    Balance-assist on $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
    Balance-assist on $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
    Balance-assist on $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
    \bottomrule
  \end{tabular}
\end{table}
